---
title: "Practical Machine Learning Project"
author: "GF"
date: "October 24, 2015"
output: html_document
---

Course Project Report - Practical Machine Learning Course

Background Introduction
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement ??? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data Sources
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project comes from this original source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

Format of Submission
The goal of your project is to predict the manner in which they did the exercise. This is the “classe” variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2.  You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.



##R Project Code

### Install R libraries
First make sure that the appropriate R libraries are installed to your working environment.

```{r, echo=TRUE}
library(ggplot2)
library(lattice)
library(caret)
library(RGtk2)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
```

###Load and preprocess data
We first donwload the source data to a local directory. While reading the we use the argumments ‘#DIV/0!’ and NA to identify erronenous data.  

```{r, echo=TRUE}
url_train <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
url_test <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
rawdata_train <- read.csv(url(url_train), na.strings=c("NA","#DIV/0!",""))
rawdata_test <- read.csv(url(url_test), na.strings=c("NA","#DIV/0!",""))
```

###Partition data set and prepare for cross-validation
We now partition the training set using a 60/40 split, meaning 60% of the data partition will used in the training sample and 40% for cross validation. We set a
seed for reproducibility.

```{r, echo=TRUE}
set.seed(12345)
splitdata_train <- createDataPartition(y=rawdata_train$classe, p=0.6, list=FALSE)
traindata <- rawdata_train[splitdata_train, ]
testdata <- rawdata_train[-splitdata_train, ]
dim(traindata)
dim(testdata)
```

###Perform data cleaning 

Remove near zero variance predictors

```{r, echo=TRUE}
#diagnose predictors that are zero variance predictors
zerovardata <- nearZeroVar(traindata, saveMetrics=TRUE)

# Create dataset without zero variance predictors
zervarnames <- names(traindata) %in% c("new_window","kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
traindata <- traindata[!zervarnames]

#Removing variable that is not in proper numeric formt for running machine learning algoritms
traindata <- traindata[c(-1)]

#Number of remaining observations 
dim(traindata)
```


Cleaning Variables with too many NAs. For Variables that have more than a 60% threshold of NA’s I’m going to leave them out:

```{r, echo=TRUE}

cleandata <- traindata #creating another subset to iterate in loop
for(i in 1:length(traindata)) { #for every column in the training dataset
        if( sum( is.na(traindata[, i] ) ) /nrow(traindata) >= .6 ) { #if n?? NAs > 60% of total observations
        for(j in 1:length(cleandata)) {
            if( length( grep(names(traindata[i]), names(cleandata)[j]) ) ==1)  { #if the columns are the same:
                cleandata <- cleandata[ , -j] #Remove that column
            }   
        } 
    }
}
#Confirm the number of observations and remove unnecessary dataset
dim(cleandata)
traindata <- cleandata
rm(cleandata)

```

Performing data transformation with rawdata_test 
```{r, echo=TRUE}
clean1 <- colnames(traindata)
clean2 <- colnames(traindata[, -58]) #classe column removed
testdata <- testdata[clean1]
rawdata_test <- rawdata_test[clean2]

#Number of observations
dim(testdata)
```

```{r, echo=TRUE}

#Number of observations
dim(rawdata_test)
```

Perform coersion to ensure data consistency.

```{r, echo=TRUE}
for (i in 1:length(rawdata_test) ) {
        for(j in 1:length(traindata)) {
        if( length( grep(names(traindata[i]), names(rawdata_test)[j]) ) ==1)  {
            class(rawdata_test[j]) <- class(traindata[i])
        }      
    }      
}

#Confirm coersion:
rawdata_test <- rbind(traindata[2, -58] , rawdata_test) 
rawdata_test <- rawdata_test[-1,]
dim(rawdata_test)

```

###Using ML algorithms for prediction: Decision Tree

```{r, echo=TRUE}
modDecTree<- rpart(classe ~ ., data=traindata, method="class")
predDecTree <- predict(modDecTree, testdata, type = "class")

```

Using confusion Matrix to test results: The accuracy of .88 is high at 95% CI, but could potentially be improved with another algorithm. We will try Random Forest to aim for a better accuracy for our model.

```{r, echo=TRUE}
confusionMatrix(predDecTree, testdata$classe)
```

###Using ML algorithms for prediction: Random Forests

```{r, echo=TRUE}
modRanForest <- randomForest(classe ~. , data=traindata)
predRanForest <- predict(modRanForest, testdata, type = "class")
```


The following confusion matrix illustrates the results  of the model:

```{r, echo=TRUE}
confusionMatrix(predRanForest, testdata$classe)
```

The results indicates that the Random Forest algorithm gives better results. This odel perfroms noticable better in
the categories across CLass A to E.

```{r, echo=TRUE}
predRanForest <- predict(modRanForest, rawdata_test, type = "class")
```

This section generates the  with predictions to submit
```{r, echo=TRUE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predRanForest)
```

